The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-08-21 sections/results.tex
Arjen's corrections in result part
(Further reworking needed)

Using climate forcing from WorldClim data, the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.

Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings. In all simulations, ice accumulates and glaciers form as a result of the artificially cooled climate. Yet the magnitude of glaciation appears very different from one climate forcing to another. Whereas the NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, the WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data. The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.

As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we run simulations using offset values ranging from 0 to 15\degC for each climate forcing.

Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(96) simulations, grouped by climate forcing. It is again visible that different forcings generally lead to very different final ice cover. Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated. Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes. Differences in regional patterns can also be noted. For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.

Final ice volume and final ice coverage is quantified in Figure~\ref{fig:ivolarea}. The NCEP/NCAR and CFSR forcing leads to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings. The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to an arbitrary target of $2\,\times10^6\,\unit{km^2}$. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

>>>>

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies (section~\ref{sec:climate}), PISM (section~\ref{sec:model}) was run using 0 to 15\degC  temperature offsets for 10~kyr to mimic glacial inception and the growth of the Cordilleran ice sheet towards its LGM configuration. Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig. \ref{fig:cool06}-\ref{fig:best}).

To contrast the six different input climatologies, we plot simulations run with a 6\degC temperature offset (~\ref{fig:cool06}). Clearly, the magnitude of glaciation differs dramatically from one climate forcing to another. In fact, differences between least extent and most extent differ by a factor above 3 and least volume and most volume by a factor above 6 (Fig. \ref{fig:ivolarea}). Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (lower panels, Fig.-\ref{fig:cool06}), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (upper panels, Fig.~\ref{fig:cool06}). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography. The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.

As illustrated in Figure \ref{fig:cool06}, spatial patterns and rates of ice sheet growth differ tremendously between the input climatologies, given a certain temperature depression. Because we performed a total of 8 simulations for each input climatology, we derived an ensemble of 96 simulations (Fig. \ref{fig:extent}). Despite temperature depressions as large as 15 °C to the input reference climatologies it is again visible that each leads to a different final ice sheet extent and volume (Fig. \ref{fig:extent} and \ref{fig:ivolarea}). Notably, for the smallest temperature offset value of 0\degC, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produce local ice caps only. In the case of WorldClim, ice is restricted to the Wrangell-St.~Elias mountains, an area presently glaciated. Several of the NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries. Differences in regional patterns can also be noted. For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.

Final ice volume and final ice coverage are shown in Figure~\ref{fig:ivolarea}. NCEP/NCAR and CFSR climatologies lead to much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings. The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.
